# R Basics

## Vectors

### Creating Vectors
```{r, eval=FALSE}
# basic vector
v <- c(2,3,8,5)

#repeated values
rep(c(1,2),5) 

#sequence
seq(4,7) #inclusive, and by default counts by 1
#shorthand
4:7

#custom length or increments
seq(0, 10, by = 2)
seq(0, 3, length.out = 7)

``` 

### Accessing Vectors
```{r, eval=FALSE}
myData <- c(2,3,8,5)

#element wise
myData[2] #selects second entry

myData[-3] #excludes third element

#can also pass in vectors to index it
myData[c(1,4)]
myData[2:4]
```


### Vector Functions

**Essential Functions**

- ```mean()```
- ```sd()```
- ```var()```
- ```max()```
- ```min()```
- ```median()```
- ```range()```
- ```quantile()```
- ```cumsum()```
- ```sum()```

**Other Functions**

- ```sort()``` - sorts a vector (alphabetically or by increasing size when numerical)
- ```rank()``` - provides the rank of each element
- ```order()``` - gives the indices of the elements in order
- ```unique()``` - returns just the unique values in the vector
- ```length()``` - total number of elements in the vector
- ```paste()``` - makes each element in the vector a string

**Particularly Interesting Functions**

- ```sample()``` - randomly sample from the elements of a vector
```{r eval=FALSE}
sample(c(3,7,9,23,45), 3, replace = FALSE)
#sample from vector, choosing n=3 without replacement
```

- ```table()``` - provide counts of the occurrence of each element
```{r eval=FALSE}
table(sample(1:6, 200, replace = TRUE))
```

- ```is.na()``` - gives a `TRUE`/`FALSE` vector as the output checking is an entry is `NA`



## Data Frames

**Structure and viewing data frames**

To show a structure of a variable, dataframe, list etc, run ```str(x)```.

```View(mydata)``` to see in spreadsheet view


### Creating Data Frames
**Manually**
```{r}
hw <- data.frame(Height = c(147, 150, 152),
                Weight = c(52.2, 53.1, 54.4))
```

**Load from ```*.csv``` file**

If in working directory,
```{r}
hw <- read.csv("hw.csv")
```
Can also use RStudio "Import Dataset" button in Environment tab (top left)

**Built in Data Sets**
Can use data frames built into R, for example:
```{r, eval=FALSE}
data("cars")
```
Run ```data()``` to see all data sets

### Accessing Data Frames
- Single column (alternative ways):
  - will collapse Data Frame column into vector
```{r, eval=FALSE}
hw$Height
hw[,1]
```
  - keeps data frame structure only keeping 1 column
```{r, eval=FALSE}
hw[,1, drop = FALSE]
```
- Single element: ```hw$Height[3]``` or ```hw[3,1]```
- Single row: ```hw[3,]``` (similar to single column)

**Full access**

```df[ r , c ]```

- r can be empty, integer between 1 and number of row, vector of integers or a vector of TRUE/FALSE of length num rows (must be exact length).
- c can be empty, integer between 1 and number of row, vector of integers, a vector of TRUE/FALSE of length num rows (must be exact length) or vector of strings with column/variable names.

**Examples** (first 3 equivalent)
```{r, eval=FALSE}
wq.red$ph
wq.red[, 9]
wq.red[, "pH"]

wq.red[, c(9,11)]
wq.red[, c("pH","alcohol")]
```

#### Advanced Query - Accessing a Subset of a Data Frame

**Examples**

- Entire column "Weight" with values >50
```{r, eval=FALSE}
hw[hw$Weight>50,]
```
- Both columns, 4 **random** rows (using ```sample()```)
```{r, eval=FALSE}
hw[sample(1:nrow(hw), 4),]
```
-**Removing NA values**
```{r, eval=FALSE}
movies[!is.na(movies$budget),] #doesn't have any NA value
```
*Note: ```!``` is the logical NOT operator and ```is.na()``` is a function that acts on a vector giving a `TRUE`/`FALSE` vector as the output*

- extra example
```{r, eval=FALSE}
wq.red[wq.red$pH>3 & wq.red$density<1, "alcohol"]
#wq.red$pH>3 is a vector of TRUE/FALSE
#wq.red$density<3 is a vector of TRUE/FALSE
#& takes logical between two vectors
```



### Interrogating Data Frames

**Data Frame Functions - information**

- ```names()``` - column names
- ```dim()``` - number of rows and columns
- ```nrow()``` - number of rows
- ```ncol()``` - number of columns
- ```head()``` - **useful for large data**, just shows top rows (can add extra parameter to specify how many rows show up)
- ```str()``` - shows details about type of data


**Data Frame Functions - interesting data**

- ```colMeans()```
- ```rowMeans()```
- ```colSums()```
- ```rowSums()```
- ```cov()``` - covariance matrix
- ```cor()``` - correlation matrix
- ```scale()``` - scales data to be centered at 0 and scaled (*both have optional arguments available*)
- ```summary()``` - gives all major statistics for each variable (column)

**Data Frame (column/row) Functions - sorting and ordering**

(*optional ```decreasing = TRUE``` argument*)

- ```sort(my.data$var)`` - sorts a variable but only outputs that column vector sorted
- ```order(my.data$var)`` - outputs a list of indices sorted

**Sorting a Data Frame by a Column**

- ```my.data[order(my.data$var),]``` - sorts whole data frame according to ```var``` column


### Manipulating Data Frames


**Creating/Adding Variables to a Data Frame**

Reference a variable that doesn't exist and just assign it to something. For example,
```{r eval=FALSE}
hw$BMI <- hw$Weight/(hw$Height/100)^2
```


#### Merging Data Frames

**```rbind(,)``` and ```cbind(,)```**

**Note: can be very error prone and often better to use tidyverse** (unless with `rbind` variables are identical and in same order or with `cbind`, observations are in same order)

- ```rbind()``` pastes rows together (above/below)
- ```cbind()``` pastes columns together (left/right)

For an example with ```rbind(,)```, if two data frames have **same column names**, ```rbind(,)``` will stack the rows to make one data frame. As an example,

```{r}
test1 <- data.frame(Col1 = c(1,5,9,2),
                    Col2 = c(6,9,8,3))
test2 <- data.frame(Col2 = c(5,9,0,1),
                    Col1 = c(4,9,3,0))

rbind(test1, test2)
```
(*notice how the columns were matched by name not order*)




### Missing Data in Data Frames

#### Importing Data with Missing Values

When using ```read.csv("mydata.csv")```, can add the additional argument ```na.strings=c(...)``` to set any strings in the vector to be replaced by ```<NA>```.

As an example,
```{r, eval=FALSE}
read.csv("carsdata.csv", na.strings = c("", "na"))
```

would import the ```cardata.csv``` file as a data frame with all strings that are empty ```""``` or ```"na"``` with the approptiate ```<NA>``` tag.

**Ignoring `NA` values - `na.rm = TRUE`**

```na.rm = TRUE``` argument ignores all ```<NA>``` values when performing the function. As an example,

```{r, eval=FALSE}
mean(mydata$var, na.rm = TRUE)
```
which is equivalent to
```{r, eval=FALSE}
mean(na.omit(mydata$var))
```

#### Remove Ovservations with ```<NA>``` Entry (```na.omit()```)

Example,
```{r, eval=FALSE}
na.omit(chickwts$weight)
```



























## Lists
```{r, eval=FALSE}
x <- list(1, "a", c(1,2,3), data.frame(a = 1:3, b = 4:6))
```

- can have a list of any type of variable
- can be good for hierarchical and tree structures, Nesting is permitted (i.e. lists can contain lists)
- variables in the list can also be named

```{r}
x <- list(bob = 1, jill = "a", jack = c(1,2,3),
          eve = data.frame(a = 1:3, b = 4:6))
x$eve$a
```

### Accessing lists
```{r eval=FALSE}
x[] #accesses an element of a list (returns a list of one element)
x[[]] #strips away one level of hierarchy, list structure is gone
x$bob #equivalent to above when items in list are named
```


## Data Types
- Numeric
- Logical (`TRUE`/`FALSE`)
- Categorical (called factors in R), could be ordered (e.g. credit rating) or could not be (e.g. eye colour)
- Date/Time
- Text or String
- Others (e.g. image, spatial, audio, video)

### Categorical Data (Factors)

#### Making Data Frames Categorical
R cannot tell the difference between factors and strings when importing data frames, so we must "tidy" them up after importing

Cleaning up a data frame (all non-numeric by default get imported as strings) by **making all strings in one variable a factor**
```{r eval=TRUE}
eyesDF <- data.frame(name=c("anne","john","charlie","sarah","max","ellie","eve"),
                     eyeColour=c("blue","green","brown","brown","blue","blue","brown"))

#changes a variable to be categorical (a factor)
eyesDF$eyeColour <- as.factor(eyesDF$eyeColour)

summary(eyesDF$eyeColour)
```

**Treat data as a factor or a text?**

- Treat as text not factor when every observation is unique (e.g. surname)
- When some text is coming up very often it may be more appropriate as a factor

#### Making Vectors Categorical

We can create a vector with factors in by creating the entries as strings and then **applying the `factor()` function**

For example,
```{r}
eye.colour <- factor(c("blue","brown","green","blue","blue","brown","green","blue","blue","green","blue","green","blue","blue","brown","green","brown","brown","green"))
summary(eye.colour)
```


**Example Using Factors**
The following data frame has a factor variable (feed)
```{r}
data("chickwts")
head(chickwts)
summary(chickwts)
```

We can filter out specific factors,
```{r eval=FALSE}
chickwts[chickwts$feed %in% c("sunflower","linseed"),]
```

**Number of levels** (different factors) and **names of levels**
```{r}

nlevels(chickwts$feed)

levels(chickwts$feed)
```


```unclass()``` gives each entry a number corresponding to a factor

```{r}
unclass(chickwts$feed)
```



## Other

**Packages**

- Installing a package: `install.packages("dplyr")`
- Loading a package: `library("dplyr")`

**Loops**
```{r, eval=FALSE}
for(i in x) {
  #i iterates over the values in vector x
}
```


**Functions**
```{r, eval=FALSE}
myFunction <- function(arg1, arg2 = 1) { 
  ...
  return(...)
}
```

**Functions acting on vectors - `sapply()`**
```{r, eval=FALSE}
sapply(X, Fun)
```
Applies function `Fun` to vector `X` element-wise

**Logical Operators**

Logical operators act element-wise on pairs of vectors (of same size) of `TRUE` and `FALSE` values. AND is `&`, OR is `|`.


