# R Basics

## Misc.

### Packages

- Installing a package: `install.packages("dplyr")`
- Loading a package: `library("dplyr")`


### Loops
```{r, eval=FALSE}
for(i in x) {
  #i iterates over the values in vector x
}
```
You cannot modify the vector `x` from within the loop.



### Functions
```{r, eval=FALSE}
myFunction <- function(arg1, arg2 = 1) { 
  ...
  return(...)
}
```

To apply a function to a vector, use
```{r, eval=FALSE}
sapply(X, Fun)
```
and `Fun` will be applied to `X` element-wise

### Logical Operators

Logical operators act element-wise on pairs of vectors (of same size) of `TRUE` and `FALSE` values. AND is `&`, OR is `|`.

## Vectors

### Creating Vectors
```{r, eval=FALSE}
# basic vector
v <- c(2,3,8,5)

#repeated values
rep(c(1,2),5) 

#sequence
seq(4,7) #inclusive, and by default counts by 1
#shorthand
4:7

#custom length or increments
seq(0, 10, by = 2)
seq(0, 3, length.out = 7)

``` 

### Accessing Vectors
```{r, eval=FALSE}
myData <- c(2,3,8,5)

#element wise
myData[2] #selects second entry

myData[-3] #excludes third element

#can also pass in vectors to index it
myData[c(1,4)]
myData[2:4]
```


### Vector Functions

**Essential Functions**

- ```mean()```
- ```sd()```
- ```var()```
- ```max()```
- ```min()```
- ```median()```
- ```range()```
- ```quantile()```
- ```cumsum()```
- ```sum()```

**Other Functions**

- ```sort()``` - sorts a vector (alphabetically or by increasing size when numerical)
- ```rank()``` - provides the rank of each element
- ```order()``` - gives the indices of the elements in order
- ```unique()``` - returns just the unique values in the vector
- ```length()``` - total number of elements in the vector
- ```paste()``` - makes each element in the vector a string

**Particularly Interesting Functions**

- ```sample()``` - randomly sample from the elements of a vector
```{r eval=FALSE}
sample(c(3,7,9,23,45), 3, replace = FALSE)
#sample from vector, choosing n=3 without replacement
```

- ```table()``` - provide counts of the occurrence of each element
```{r eval=FALSE}
table(sample(1:6, 200, replace = TRUE))
```

- ```is.na()``` - gives a `TRUE`/`FALSE` vector as the output checking is an entry is `NA`



## Lists
```{r, eval=FALSE}
x <- list(1, "a", c(1,2,3), data.frame(a = 1:3, b = 4:6))
```

- can have a list of any type of variable
- can be good for hierarchical and tree structures, Nesting is permitted (i.e. lists can contain lists)
- variables in the list can also be named

```{r}
x <- list(bob = 1, jill = "a", jack = c(1,2,3),
          eve = data.frame(a = 1:3, b = 4:6))
x$eve$a
```

### Accessing lists
```{r eval=FALSE}
x[] #accesses an element of a list (returns a list of one element)
x[[]] #strips away one level of hierarchy, list structure is gone
x$bob #equivalent to above when items in list are named
```


## Data Types
- Numeric
- Logical (`TRUE`/`FALSE`)
- Categorical (called factors in R), could be ordered (e.g. credit rating) or could not be (e.g. eye colour)
- Date/Time
- Text or String
- Others (e.g. image, spatial, audio, video)

### Categorical Data (Factors)

#### Making Data Frames Categorical
R cannot tell the difference between factors and strings when importing data frames, so we must "tidy" them up after importing

Cleaning up a data frame (all non-numeric by default get imported as strings) by **making all strings in one variable a factor**
```{r eval=TRUE}
eyesDF <- data.frame( name=c("anne","john","charlie","sarah","max","ellie","eve"),
                      eyeColour=c("blue","green","brown","brown","blue","blue","brown"))

#changes a variable to be categorical (a factor)
eyesDF$eyeColour <- as.factor(eyesDF$eyeColour)

summary(eyesDF$eyeColour)
```

**Treat data as a factor or a text?**

- Treat as text not factor when every observation is unique (e.g. surname)
- When some text is coming up very often it may be more appropriate as a factor

#### Making Vectors Categorical

We can create a vector with factors in by creating the entries as strings and then **applying the `factor()` function**

For example,
```{r}
eye.colour <- factor(c("blue","brown","green","blue","blue","brown","green","blue","blue","green","blue","green","blue","blue","brown","green","brown","brown","green"))
summary(eye.colour)
```


**Example Using Factors**
The following data frame has a factor variable (feed)
```{r}
data("chickwts")
head(chickwts)
summary(chickwts)
```

We can filter out specific factors,
```{r eval=FALSE}
chickwts[chickwts$feed %in% c("sunflower","linseed"),]
```

**Number of levels** (different factors) and **names of levels**
```{r}

nlevels(chickwts$feed)

levels(chickwts$feed)
```


```unclass()``` gives each entry a number corresponding to a factor

```{r}
unclass(chickwts$feed)
```

