# Plotting

## Plotting in Base R

(useful for exploration of data)

### Core Plot Function

Plots data ```x``` against data ```y```. (*if only ```x``` is supplied, the indices will be plotted against their values*)

```{r, eval=FALSE}
plot(x,y, ...)
```

**Optional Arguments**

- ```col``` - colour of points (can use RGB or colour name as a string; can be vector for each point)
- ```pch``` - plotting symbol (cross, circle etc), an integer
  - [Plotting symbol reference](https://cpb-us-e1.wpmucdn.com/sites.ucsc.edu/dist/d/276/files/2015/10/symbols.jpg)
- ```xlab``` and ```ylab``` - labels
- ```xlim``` and ```ylim``` - limits in the form of a 2-vector (e.g. ```xlim = c(20,100)``` restricts x from 20 to 100)
- ```main``` - Plot Title
- ```type```
  - ```"p"``` - points (default)
  - ```"l"``` - line connecting observations
  - ```"b"``` - both points and lines


### Other Plot Functions

- ```hist()``` - Histogram
- ```boxplot()``` - Boxplot
- ```barplot()``` - Categorical Bar Charts (use ```table``` to get summary)


*Note: you can store plots inside of variables*
```{r, eval=TRUE, fig.show='hide'}

data("diamonds", package = "ggplot2")

hgram <- hist(diamonds$price, freq = FALSE)

str(hgram)
```

### Adding to Plots

Each ```plot()``` function creates a new plot. To add to an existing plot use,

- ```points()``` - adds a plot of points to an existing plot
- ```lines()``` - shorthand for ```points(x, y, type="l")```
- ```abline()``` - adds a $y=mx+c$ line directly

#### Fitting Lines to Plots

(see linear regression)

- ```lm()``` - **fits a straight line**, pass inside of ```abline()```
- ```lowess()``` - **fits a smooth line**, pass inside of ```lines()``` (```f``` argument controls smoothness)
- ```density()``` - fits a smooth continuous version of a histogram

**Example**
```{r eval=TRUE}
data("diamonds", package = "ggplot2")
plot(diamonds$carat, diamonds$price, pch = 20)


abline(lm(price ~ carat, diamonds),
       col = "red")
lines(lowess(diamonds$carat, diamonds$price, f = 0.05),
      col = "green")


hist(diamonds$price, freq = FALSE)
lines(density(diamonds$price), col = "red")
```


### Multiple Plots
(often better to just use ggplot2)

To get a grid of all pairwise scatter plots, use ```pairs()```
```{r eval=FALSE}
pairs(mtcars)

pairs(mtcars[,1:4])
```

You can also **manually set the grid size** (using ```par(mfrow = c(n,m))```) and then populate each grid slot one by one by calling slots
```{r eval=FALSE}
par(mfrow = c(2,1))
plot(diamonds$carat, diamonds$price)
boxplot(diamonds$carat)
par(mfrow = c(1,1)) # <- need this to reset to a single plot!
```

To reset the plotting window to default use `dev.off()`.

## Plotting in ```ggplot2```

(useful for presentation of data)

Loading ```ggplot2```,
```{r, eval=FALSE}
# Either ...
library("tidyverse")
# for all tidyverse packages
# OR, for just plotting
library("ggplot2")
```

### Main Structure

**Starting a plot**

Every plot starts with the function ```ggplot()``` with the optional arguments:

- data - to specify the data frame containing the variables we later reference
- ```aes()``` - mapping to specify what variables map to the x axis, y axis, colour legend, etc

For example,
```{r echo=FALSE, message=FALSE}
library("tidyverse")
data("diamonds", package = "ggplot2")
```
```{r, out.width='40%', fig.align = 'center'}
ggplot(diamonds, aes(x = carat, y = price))
```

Axis are labelled and scaled but nothing is plotted yet (as we have not called a "Geom").


**Geoms**

A ```geom_``` will add a layer to the plot. Examples of Geoms:

- ```geom_point()``` - most basic, plots ```x``` against ```y``` as scatter plot
- ```geom_line()```
- ```geom_smooth()``` - smoothed curve (defaukt method is "gam", can also use "lm")
- ```geom_bar()``` - barchart (1 variable and counts)
- ```geom_col()``` - barchart (2 variables)
- ```geom_boxplot()``` - boxplot

More unusual ones,

- ```geom_hex()```
- ```geom_polygon()```

**```aes()```**

If you want to specify the ```x``` and ```y``` variables, colour by a property, group by a property, change the point size based on a property etc then you put that information into the ```aes()```.

The ```aes(...)``` that goes into the original ```ggplot(aes())``` will be inherited by all plots unless overridden. The ```aes(...)``` that goes into a particular geom, ```geom_...(aes())``` only applies to that geom.

**Labels**

```xlab("X-axis Label")```, ```ylab("Y-axis Label")``` and ```ggtitle("Title")``` can also be added to the plot in the same way as Geoms.

Alternativily, use ```+ labs(title="Title", x="X-axis", y="Y-axis")```


### Updating a Plot (Plots in Variables)

```{r, eval=FALSE}
data("mtcars")
p <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point()

p + geom_smooth()

p + geom_smooth(method = "lm")

p + scale_y_log10() + scale_x_log10() +
  geom_smooth(method = "lm")

p + scale_y_log10() + scale_x_log10() +
  geom_smooth(method = "lm") +
  geom_vline(xintercept = 100)
```

Here, ```p``` stores the basic plot and each time, something different is added to it for a new plot, but without updating ```p```.

### Faceting

Faceting enables splitting your data into multiple plots according to a categorical variable.

- ```facet_wrap()``` - a single variable split
  - formula notation to indicate splitting variable ```~ var```
  - optionally specify number of rows
- ```facet_grid()``` - two variable split
  - formula indicating both splitting variables rows_var ~ cols_var

formula indicating both splitting variables rows_var ~ cols_var

For example,

```{r, echo=FALSE, message=FALSE}
data("mtcars")
```
```{r, out.width='40%', fig.align = 'center'}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  facet_wrap(~ gear) +
  geom_point()

ggplot(mtcars, aes(x = hp, y = mpg)) +
  facet_grid(cyl ~ gear) +
  geom_point()
```

### Examples

```{r, message=FALSE, echo=FALSE}
data("mpg", package = "ggplot2")
```
```{r, message=FALSE, out.width='40%', fig.align = 'center'}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut), size = 0.2) +
  geom_smooth(aes(colour = cut)) +
  xlab("Number of carats") + ylab("Price in $")

ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(colour = class))

ggplot(mpg, aes(x=displ, y=hwy)) +
  facet_wrap(~class) +
  geom_point() +
  geom_point(aes(y=cty), colour="red") + #aes() doesn't hold all information!!!
  ylab("Fuel efficiency")


ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point(aes(colour=drv)) +
  geom_smooth(colour="black") +
  geom_smooth(aes(colour=drv))


ggplot(mpg, aes(x=class)) +
  geom_bar(aes(fill=drv))
```