# Splines

See final subsection on comparing splines for best way to plot this kind of data.

## Step Functions

With `Boston` data, following on from polynomial regression section.

```{r echo=FALSE, warning=FALSE}
library("MASS")
y = Boston$medv
x = Boston$lstat
y.lab = 'Median Property Value'
x.lab = 'Lower Status (%)'
```

To use a step function model, use `cut(x, n)` function which automatically assigns break points. Note that `n` specifies the number of intervals, **not the number of break points (cuts)**.

```{r}
step3 = lm(y ~ cut(x, 3))
```

```{r echo=FALSE}
step2 = lm(y ~ cut(x, 2))
#step3 = lm(y ~ cut(x, 3))
step4 = lm(y ~ cut(x, 4))
step5 = lm(y ~ cut(x, 5))
```

Then following generic analysis for plotting:

```{r eval=FALSE}
pred3 = predict(step3, newdata = list(x = sort(x)), se = TRUE)

se.bands3 = cbind(pred3$fit + 2*pred3$se.fit, pred3$fit-2*pred3$se.fit)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab,
     main = "2 cutpoints", bty = 'l')
lines(sort(x), pred3$fit, lwd = 2, col = "darkviolet")
matlines(sort(x), se.bands3, lwd = 1.4, col = "darkviolet", lty = 3)
```

We plot below as if we did it for 2,3,4 and 5 intervals:

```{r echo=FALSE}
pred2 = predict(step2, newdata = list(x = sort(x)), se = TRUE)
pred3 = predict(step3, newdata = list(x = sort(x)), se = TRUE)
pred4 = predict(step4, newdata = list(x = sort(x)), se = TRUE)
pred5 = predict(step5, newdata = list(x = sort(x)), se = TRUE)

se.bands2 = cbind(pred2$fit + 2*pred2$se.fit, pred2$fit-2*pred2$se.fit)
se.bands3 = cbind(pred3$fit + 2*pred3$se.fit, pred3$fit-2*pred3$se.fit)
se.bands4 = cbind(pred4$fit + 2*pred4$se.fit, pred4$fit-2*pred4$se.fit)
se.bands5 = cbind(pred5$fit + 2*pred5$se.fit, pred5$fit-2*pred5$se.fit)

par(mfrow = c(2,2))

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab,
     main = "1 cutpoint", bty = 'l')
lines(sort(x), pred2$fit, lwd = 2, col = "red")
matlines(sort(x), se.bands2, lwd = 1.4, col = "red", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab,
     main = "2 cutpoints", bty = 'l')
lines(sort(x), pred3$fit, lwd = 2, col = "darkviolet")
matlines(sort(x), se.bands3, lwd = 1.4, col = "darkviolet", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab,
     main = "3 cutpoints", bty = 'l')
lines(sort(x), pred4$fit, lwd = 2, col = "blue")
matlines(sort(x), se.bands4, lwd = 1.4, col = "blue", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab,
     main = "4 cutpoints", bty = 'l')
lines(sort(x), pred5$fit, lwd = 2, col = "black")
matlines(sort(x), se.bands5, lwd = 1.4, col = "black", lty = 3)
```



## Regression Splines

Using `Boston` data again with `x` and `y` defined as variables for convenience with labels `y.lab` and `x.lab` done as well.

```{r echo=FALSE}
library(MASS)
y = Boston$medv
x = Boston$lstat
y.lab = 'Median Property Value'
x.lab = 'Lower Status (%)'
```

For regression splines, we use the `splines` package.

```{r warning=FALSE, error=FALSE}
library(splines)
```

Cuts for regression splines can be specified or done automatically. One option is to specify them using quartiles: `cuts = summary(x)[c(2, 3, 5)]`.


Since we use data points $x$ values as points to plot the splines, we must also sort them: `sort.x = sort(x)`.

```{r echo=FALSE}
cuts = summary(x)[c(2, 3, 5)] 
sort.x = sort(x)
```

**Regression Spline Model and SE Bands**

We use `bs()` function from `splines` and can either specify the knots or specify the degrees of freedom:

```{r}
spline1 = lm(y ~ bs(x, degree = 1, knots = cuts))
pred1 = predict(spline1, newdata = list(x = sort.x), se = TRUE)
se.bands1 = cbind(pred1$fit + 2 * pred1$se.fit, pred1$fit - 2 * pred1$se.fit)
```

```{r}
spline1df = lm(y ~ bs(x, degree = 1, df = 5))
pred1df = predict(spline1df, newdata = list(x = sort.x), se = TRUE)
se.bands1df = cbind( pred1df$fit + 2 * pred1df$se.fit, 
                     pred1df$fit - 2 * pred1df$se.fit )
```

**Plot**

```{r}
par(mfrow = c(1, 2))

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Linear Spline (with knots)", bty = 'l')
lines(sort.x, pred1$fit, lwd = 2, col = "red")
matlines(sort.x, se.bands1, lwd = 2, col = "red", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Linear Spline (with df)", bty = 'l')
lines(sort.x, pred1df$fit, lwd = 2, col = "darkred")
matlines(sort.x, se.bands1df, lwd = 2, col = "red", lty = 3)
```

**For Higher Degrees**

```{r}
spline2 = lm(y ~ bs(x, degree = 2, df = 6))
pred2 = predict(spline2, newdata = list(x = sort.x), se = TRUE)
se.bands2 = cbind(pred2$fit + 2 * pred2$se.fit, pred2$fit - 2 * pred2$se.fit)

spline3 = lm(y ~ bs(x, degree = 3, df = 7))
pred3 = predict(spline3, newdata = list(x = sort.x), se = TRUE)
se.bands3 = cbind(pred3$fit + 2 * pred3$se.fit, pred3$fit - 2 * pred3$se.fit)

par(mfrow = c(1,3))
plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Linear Spline", bty = 'l')
lines(sort.x, pred1$fit, lwd = 2, col = "darkred")
matlines(sort.x, se.bands1, lwd = 2, col = "darkred", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Quadratic Spline", bty = 'l')
lines(sort.x, pred2$fit, lwd = 2, col = "darkgreen")
matlines(sort.x, se.bands2, lwd = 2, col = "darkgreen", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Cubic Spline", bty = 'l')
lines(sort.x, pred3$fit, lwd = 2, col = "darkblue")
matlines(sort.x, se.bands3, lwd = 2, col = "darkblue", lty = 3)
```



## Natural Splines

Adds the extra constraints of continuity in first $d-1$ derivatives. We use the `ns()` function. Again, we use `library("splines")` package.

```{r eval=FALSE}
spline.ns3 = lm(y ~ ns(x, df = 3))
pred.ns3 = predict(spline.ns3, newdata = list(x = sort.x), se = TRUE)
se.bands.ns3 = cbind(pred.ns3$fit + 2 * pred.ns3$se.fit, 
                     pred.ns3$fit - 2 * pred.ns3$se.fit)
```

```{r echo=FALSE}
spline.ns1 = lm(y ~ ns(x, df = 1))
pred.ns1 = predict(spline.ns1, newdata = list(x = sort.x), se = TRUE)
se.bands.ns1 = cbind(pred.ns1$fit + 2 * pred.ns1$se.fit, 
                     pred.ns1$fit - 2 * pred.ns1$se.fit)

spline.ns2 = lm(y ~ ns(x, df = 2))
pred.ns2 = predict(spline.ns2, newdata = list(x = sort.x), se = TRUE)
se.bands.ns2 = cbind(pred.ns2$fit + 2 * pred.ns2$se.fit, 
                     pred.ns2$fit - 2 * pred.ns2$se.fit)

spline.ns3 = lm(y ~ ns(x, df = 3))
pred.ns3 = predict(spline.ns3, newdata = list(x = sort.x), se = TRUE)
se.bands.ns3 = cbind(pred.ns3$fit + 2 * pred.ns3$se.fit, 
                     pred.ns3$fit - 2 * pred.ns3$se.fit)

spline.ns4 = lm(y ~ ns(x, df = 4))
pred.ns4 = predict(spline.ns4, newdata = list(x = sort.x), se = TRUE)
se.bands.ns4 = cbind(pred.ns4$fit + 2 * pred.ns4$se.fit, 
                     pred.ns4$fit - 2 * pred.ns4$se.fit)
```


**Plotting**

Assuming we have made similarly `spline.ns1`, `spline.ns2`, `spline.ns3` and `spline.ns4`.

```{r}
par(mfrow = c(2, 2))

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Natural Spline (1 df)", bty = 'l')
lines(sort.x, pred.ns1$fit, lwd = 2, col = "darkred")
matlines(sort.x, se.bands.ns1, lwd = 2, col = "darkred", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Natural Spline (2 df)", bty = 'l')
lines(sort.x, pred.ns2$fit, lwd = 2, col = "darkgreen")
matlines(sort.x, se.bands.ns2, lwd = 2, col = "darkgreen", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Natural Spline (3 df)", bty = 'l')
lines(sort.x, pred.ns3$fit, lwd = 2, col = "darkblue")
matlines(sort.x, se.bands.ns3, lwd = 2, col = "darkblue", lty = 3)

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Natural Spline (4 df)", bty = 'l')
lines(sort.x, pred.ns4$fit, lwd = 2, col = "brown")
matlines(sort.x, se.bands.ns4, lwd = 2, col = "brown", lty = 3)
```

**Practical 4 Example (Using Knots)**

```{r eval=FALSE}
ns = lm(y ~ ns(x, knots = cuts))
pred.ns = predict(ns, newdata = list(x = grid), se = TRUE)
se.bands.ns = cbind(pred.ns$fit + 2 * pred.ns$se.fit, 
                     pred.ns$fit - 2 * pred.ns$se.fit)
```


## Smoothing Splines


For smoothing splines we use `smoothing.splines()` instead of `lm()`. Use the `df` parameter to specify effective degrees of freedom or the `cv = TRUE` parameter to use (ordinary leave-one-out) cross validation instead (to choose a $\lambda$). Can also directly specify `lambda` or `spar` (a smoothing parameter).

**Notes:** Specifying $\lambda$ with effective degrees of freedom (`df = ...`) still uses cross-validation.

```{r warning=FALSE}
smooth1 = smooth.spline(x, y, df = 3) #choose lambda using effective degrees of freedom
smooth2 = smooth.spline(x, y, cv = TRUE) #choose lambda via cross validation

par(mfrow = c(1,2))
plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Smoothing Spline (3 df)", bty = 'l')
lines(smooth1, lwd = 2, col = "brown")

plot(x, y, cex.lab = 1.1, col="darkgrey", xlab = x.lab, ylab = y.lab, 
     main = "Smoothing Spline (CV)", bty = 'l')
lines(smooth2, lwd = 2, col = "darkorange")
```



## Comparing Splines

Below compares different splines on wage data. We load packages `library("ISLR")` for data and `library("splines")` to use regression and natural splines.

```{r echo=FALSE}
library("ISLR") # Package for data
library("splines") # Required for regression and natural splines.
```

**Setup for plotting curves**
```{r}
agelims = range(Wage$age)
age.grid = seq(from=agelims[1], to=agelims[2])
```

**Taking a sample of data of size $n$**
```{r}
set.seed(1)
# Number of data points - change this to investigate small (50), 
# medium (200) and large (1000) samples.
n <- 200

ind = sample(1:3000, n)
Wage1 = Wage[ind,]  # Label subset of data as Wage1.
```

**Cubic Spline**
```{r warning=FALSE}
fitbs = lm(wage~bs(age, degree = 3, knots = c(30,40,60)), data = Wage1)
predbs = predict(fitbs, newdata = list(age = age.grid), se = T)
```

**Natural Spline**
```{r}
fitns = lm(wage~ns(age, knots = c(30,40,60)), data = Wage1)
predns = predict(fitns, newdata = list(age = age.grid), se = T)
```

**Smoothing Spline**
```{r warning=FALSE}
fitss = smooth.spline(Wage1$age, Wage1$wage, cv = TRUE) 
```

**Generate the Plots**
```{r}
par(mfrow=c(1,3))

# Cubic Spline
plot(Wage1$age, Wage1$wage, col = "darkgray", pch = 19, 
     main = 'Cubic spline', bty = 'l', 
     xlab = 'age', ylab = 'wage', cex.lab = 1.4)
lines(age.grid, predbs$fit, lwd = 2, col = 'darkgreen')
abline(v = c(30,40,60), lty = 'dashed')

# Natural Spline
plot(Wage1$age, Wage1$wage, col = "darkgray", pch = 19, 
     main = 'Natural cubic spline', bty = 'l', 
     xlab = 'age', ylab = 'wage', cex.lab = 1.4)
lines(age.grid, predns$fit, lwd = 2, col = 'darkviolet')
abline(v = c(30,40,60), lty = 'dashed')

# Smoothing Spline
plot(Wage1$age, Wage1$wage, col = "darkgray", pch = 19, 
     main = 'Smoothing spline', bty = 'l', 
     xlab = 'age', ylab = 'wage', cex.lab = 1.4)
lines(fitss, lwd = 2, col = 'brown')
```




**Practical 4 Example**

Modelling `medv ~ indus` using natural and smoothing splines.

Basic Setup:
```{r}
library("MASS")
library(splines)

data("Boston")

y = Boston$medv
x = Boston$indus
y.lab = "Median House Value ($1000s)"
x.lab = "Proportion of non-retail business acres per town"
```

```{r}
cuts <- summary(x)[c(2,3,5)] #For knots
grid = seq(min(x), max(x), length.out = 500) #For plotting curves
```

Cubic Spline:
```{r}
bs = lm(y ~ bs(x, knots = cuts))
pred.bs = predict(bs, newdata = list(x = grid), se = TRUE)
se.bands.bs = cbind(pred.bs$fit + 2 * pred.bs$se.fit, 
                    pred.bs$fit - 2 * pred.bs$se.fit)
```

Natural Cubic Spline (Knots):
```{r}
nsk = lm(y ~ ns(x, knots = cuts))
pred.nsk = predict(nsk, newdata = list(x = grid), se = TRUE)
se.bands.nsk = cbind(pred.nsk$fit + 2 * pred.nsk$se.fit, 
                     pred.nsk$fit - 2 * pred.nsk$se.fit)
```

Natural Cubic Spline (2 df):
```{r}
ns2 = lm(y ~ ns(x, df = 2))
pred.ns2 = predict(ns2, newdata = list(x = grid), se = TRUE)
se.bands.ns2 = cbind(pred.ns2$fit + 2 * pred.ns2$se.fit, 
                     pred.ns2$fit - 2 * pred.ns2$se.fit)
```

Natural Cubic Spline (3 df):
```{r}
ns3 = lm(y ~ ns(x, df = 3))
pred.ns3 = predict(ns3, newdata = list(x = grid), se = TRUE)
se.bands.ns3 = cbind(pred.ns3$fit + 2 * pred.ns3$se.fit, 
                     pred.ns3$fit - 2 * pred.ns3$se.fit)
```

Smoothing Spline (3 effective df):
```{r}
smooth = smooth.spline(x, y, df = 3)
```

Plot:
```{r}
par(mfrow = c(2,3))

plot(x, y, col="darkgrey", xlab = x.lab, ylab = y.lab, main = "Cubic Spline", bty = 'l', pch = 16)
lines(grid, pred.bs$fit, lwd = 2, col = 1)
matlines(grid, se.bands.bs, lwd = 2, col = 1, lty = 3)

plot(x, y, col="darkgrey", xlab = x.lab, ylab = y.lab, main = "Natural Cubic Spline", bty = 'l', pch = 16)
lines(grid, pred.nsk$fit, lwd = 2, col = 2)
matlines(grid, se.bands.nsk, lwd = 2, col = 2, lty = 3)

plot(x, y, col="darkgrey", xlab = x.lab, ylab = y.lab, main = "Natural Spline (2 df)", bty = 'l', pch = 16)
lines(grid, pred.ns2$fit, lwd = 2, col = 3)
matlines(grid, se.bands.ns2, lwd = 2, col = 3, lty = 3)

plot(x, y, col="darkgrey", xlab = x.lab, ylab = y.lab, main = "Natural Spline (3 df)", bty = 'l', pch = 16)
lines(grid, pred.ns3$fit, lwd = 2, col = 4)
matlines(grid, se.bands.ns3, lwd = 2, col = 4, lty = 3)

plot(x, y, col="darkgrey", xlab = x.lab, ylab = y.lab, main = "Smoothing Spline (3 Effective df)", bty = 'l', pch = 16)
lines(smooth, lwd = 2, col = 6)
```