# GAMs

We use the `library(gam)` package. Below is an example using the `Boston` data again.

```{r echo=FALSE, warning=FALSE}
library(MASS)
y = Boston$medv
x = Boston$lstat
y.lab = 'Median Property Value'
x.lab = 'Lower Status (%)'
```

```{r warning=FALSE, message=FALSE}
library(gam)

names(Boston)
```

In this example we build a GAM using a cubic spline with 3 degrees of freedom for `lstat`, a smoothing spline with 3 degrees of freedom for `indus` and a simple linear model for variable `chas`. But first we make `chas` a categorical variable.

```{r}
Boston1 = Boston
Boston1$chas = factor(Boston1$chas)
```

**GAM Function**

We use the `gam()` function which has the syntax `gam( response ~ predictors + ..., data = data)`. The way to specify what the model of each predictor is is similar to `lm()`.

- `bs()` is used for a regression spline
- `ns()` is used for natural splines
- `s()` is used for smoothing splines (this is different from `lm` and how smoothing splines are done normally).

```{r}
gam = gam( medv ~ bs(lstat, degree = 3, df = 5) + s(indus, df = 5) + chas, data = Boston1 )
```

**Plotting**

We have the option to include standard error error bars or not.

```{r}
par( mfrow = c(1,3) )
plot( gam,  se = TRUE, col = "blue" )
```

**Prediction**

Prediction works the same as with `lm()` for multiple linear regression models.

```{r}
preds <- predict( gam, newdata = data.frame( chas = "0", indus = 3, lstat = 5 )  )
preds
```