# Logistic Regression

For estimating probability of a qualitative variable. See MLLN Notes for exploratory plots and motivation.

## Simple Logistic Regression

We use the `Default` data from `library("ISLR2")` package. We also load `library(scales)` package for transparency with colours.

```{r echo=FALSE, warning=FALSE}
library(scales)
library("ISLR2")
```

We use the generalised linear model function `glm()` with `family = "binomial"`.

```{r}
data(Default)

glm.fit = glm( as.numeric(Default$default=="Yes") ~ balance, data = Default, family = "binomial")
summary(glm.fit)
summary(glm.fit$fitted.values)
```

**Plot**

```{r}
#Raw Data 
plot(Default$balance, as.numeric(Default$default=="Yes"),col="red",xlab="balance",ylab="default")

#Fitted Values
points(glm.fit$data$balance,glm.fit$fitted.values, col = "black", pch = 4)

#Model Curve
curve(predict(glm.fit,data.frame(balance = x),type="resp"),col="blue",lwd=2,add=TRUE)
```


**Predicting $\hat{p}(X)$**

```{r}
X = 1300
p.hat = exp(sum(glm.fit$coefficients*c(1,X)))/(1+exp(sum(glm.fit$coefficients*c(1,X))))
p.hat
```

## Multiple Logistic Regression

**Pairs Plot (Factor Variable Coloured)**

```{r echo=FALSE}
admit <- read.csv("https://www.maths.dur.ac.uk/users/hailiang.du/data/admit.csv")
```

```{r}
pairs(admit[,2:4], col=admit[,1]+2, pch=16)
```

**Making a Variable a Factor**
```{r}
admit$rank = factor(admit$rank)
```

**Practical 4 Example**

```{r}
glm.fit = glm( admit ~ ., data = admit, family = "binomial")
summary(glm.fit)
```



## Inference

**Calculating Probability**

Use `predict()` function with `type = "response"` or `type = "resp"`. Do not include any data to calculate predictions for training data (used to plot the curve)

```{r}
glm.probs = predict(glm.fit, type = "response")
```

**Single Prediction/Multiple New Prediction**

```{r}
#Data For Prediction
testdata = data.frame(gre = 380, gpa = 3.61, rank = 3)
testdata$rank = factor(testdata$rank)

#Single Prediction
predict(glm.fit, testdata, type="resp")
```

**Predictions**

Round probability to get a prediction.

```{r}
#Prediction
glm.pred = round(glm.probs)

#Confusion Table
table(glm.pred, admit$admit)

#Ratio of correct predictions on test data
mean(glm.pred == admit$admit)
```
