# Tree Models

```{r echo=FALSE}
#Synthetic Surrogate Data
set.seed(212)
data_surrogate <- data.frame(x = c(runif(200, 0, 0.4), runif(400, 0.4, 0.65), runif(300, 0.65, 1)),
                             y = c(rnorm(200, 1.5), rnorm(400, 0), rnorm(300, 2)))
```

## Building Tree Models

```{r warning=FALSE}
library("tree")
```

Use `tree()` function from `library("tree")`.

```{r}
tree_fit = tree(y ~ x, data_surrogate)
#summary(tree_fit)
#tree_fit
```

### Plotting Trees {-}

If `pretty = 0` then the level names of a factor split attributes are used unchanged.

```{r}
plot(tree_fit)
text(tree_fit, pretty=0)
```


### Predictions {-}

Example from practical demonstration with synthetic surrogate data.

```{r}
set.seed(347)
data_surrogate_test <- data.frame(x = c(runif(200, 0, 0.4), runif(400, 0.4, 0.65), runif(300, 0.65, 1)),
                                  y = c(rnorm(200, 1.5),    rnorm(400, 0),         rnorm(300, 2)))
```

Use `predict()` function. The tree model gives the mean in that terminal node group as a prediction.

```{r}
tree_pred = predict(tree_fit, data_surrogate_test)
```

**Plotting Predictions**
```{r}
plot(x = data_surrogate_test$x[1:200], y = data_surrogate_test$y[1:200],
     col = 'blue', xlab = "X", ylab = "Y", pch = 19, xlim = c(0,1), ylim = c(-2,4))
points(x = data_surrogate_test$x[201:600], y = data_surrogate_test$y[201:600], col = 'red', pch = 19)
points(x = data_surrogate_test$x[601:900], y = data_surrogate_test$y[601:900], col = 'green', pch = 19)

points(x=data_surrogate_test$x,y=tree_pred,col='black',pch=15)
```

**MSE**
```{r}
pred_mse=mean((data_surrogate_test$y-tree_pred)^2)
pred_mse
```



### Pruning Trees {-}

```{r echo=FALSE}
set.seed(739)
data_surrogate_s <- data.frame(x = c(runif(20, 0, 0.4), runif(40, 0.4, 0.65), runif(30, 0.65, 1)),
                               y = c(rnorm(20, 1.5),    rnorm(40, 0),         rnorm(30, 2)))
tree_fit_s = tree(y~x,data_surrogate_s)
tree_pred_s = predict(tree_fit_s,data_surrogate_test)
pred_mse=mean((data_surrogate_test$y-tree_pred_s)^2)
```


Prune tree with `prune.tree()` command, but can first plot the `deviance` for different size trees using `cv.tree()` with parameter `FUN = prune.tree`.

```{r}
tree_cv_prune = cv.tree(tree_fit_s, FUN = prune.tree)
#tree_cv_prune
plot(tree_cv_prune)
```
