# Misc Exam Notes (Temporary)

Make sure to look at if the examples i'm basing my solution off is for classification or regression. For binary classification problems, use `type = "response"` in `predict()` function.

** If, Else Function (`ifelse(condition, condition_true, condition_false)`)**

Example: `status_pred=ifelse(pred.boost<=0.5,"bad","good")`

## Plotting

**Basic Plot**
```{r echo=FALSE, warning=FALSE}
library("faraway")
data("seatpos")
y = seatpos$hipcenter
x = seatpos$Ht
```

```{r}
plot(x, y, col = "darkgrey", xlab = "X Label", ylab = "Y Label", main = "Title", bty = 'l', pch = 16)
```

**Plotting with a `grid`**

Use instead of `sort(x)`
```{r}
grid = seq(min(x), max(x), length.out = 250) 
```

## Generating Synthetic Data

Use `rnorm`, `runif` etc functions to generate random data.

**Trees Example - Surrogate Data**

```{r}
set.seed(212)
data_surrogate <- data.frame(x = c(runif(200, 0, 0.4), runif(400, 0.4, 0.65), runif(300, 0.65, 1)),
                             y = c(rnorm(200, 1.5), rnorm(400, 0), rnorm(300, 2)))
```

```{r}
plot(x=data_surrogate$x[1:200], y=data_surrogate$y[1:200], col='blue',
     xlab="x", ylab="y", pch=16, xlim=c(0,1),ylim=c(-2,4), bty = 'l', main = "Synthetic Surrogate Data")
points(x=data_surrogate$x[201:600], y=data_surrogate$y[201:600], col='red', pch=19)
points(x=data_surrogate$x[601:900], y=data_surrogate$y[601:900], col='green', pch=19)
```

## `summary`

Make sure to read the summary of a model to answer any questions.